\huge UNet-HCRF Integration for EEG Pattern Recognition

### 2024-12-12 manuscript/main.tex
edit experiment

%% keep it clean; only include those you need
\usepackage{amsfonts,amsmath,amssymb}
\usepackage{bbm}
\usepackage{graphicx}
\graphicspath{{./}{../image/}}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage[numbers]{natbib}
\usepackage{xcolor}

%% notations; only define those you use frequently
\newcommand{\EE}{{\mathbb{E}}}
\newcommand{\Fb}{\mathbf{F}}
\newcommand{\hb}{\mathbf{h}}
\newcommand{\Hc}{\mathcal{H}}
\newcommand{\Ib}{\mathbf{I}}
\newcommand{\Lc}{\mathcal{L}}
\newcommand{\PP}{{\mathbb{P}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Oc}{\mathcal{O}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\Ocal}{\mathcal{O}}
\newcommand{\RR}{{\mathbb{R}}}

@@

@@

>>>>

%% keep it clean; only include those you need
\usepackage{amsfonts,amsmath,amssymb}
\usepackage{bbm}
\usepackage[inline]{enumitem}
\usepackage{graphicx}
\graphicspath{{./}{../image/}}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage[numbers]{natbib}
\usepackage{subcaption}
\usepackage{xcolor}

%% notations; only define those you use frequently
\newcommand{\EE}{{\mathbb{E}}}
\newcommand{\Fb}{\mathbf{F}}
\newcommand{\hb}{\mathbf{h}}
\newcommand{\Hc}{\mathcal{H}}
\newcommand{\Ib}{\mathbf{I}}
\newcommand{\Lc}{\mathcal{L}}
\newcommand{\PP}{{\mathbb{P}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Oc}{\mathcal{O}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\Ocal}{\mathcal{O}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\XX}{{\mathbb{X}}}
\newcommand{\YY}{{\mathbb{Y}}}

@@

\let\proglang=\textsf
\let\pkg=\texttt

@@

\section{Experiment Details}

\subsection{Dataset Description}

The dataset used in this research is the Harmful Brain Activity Classification
(HMS) dataset~\citep{jing2023development}, provided by Harvard Medical School
for research purposes. This dataset comprises EEG recordings and their
corresponding spectrograms, segmented into 50-second-long samples derived from
10-minute windows of brain activity.
The longer 10-minute window allows for the model to pick up on further insights
regarding the nature of the brain activity. Each EEG sample, representing 50
seconds of brain activity measured by 19 scalp electrodes, is paired with a
matching spectrogram.
The EEG samples were labeled by expert annotators from Critical Care EEG
Monitoring Research Consortium (CCEMRC) into one of six types of brain
activity: seizures, generalized periodic discharges (GPD), lateralized
periodic discharges (LPD), lateralized rhythmic delta activity (LRDA),
generalized rhythmic delta activity (GRDA), and other.

Expert agreement was high for some segments; nevertheless, instances of
disagreement were observed. The segments were categorized as follows:
\begin{enumerate*}[label = (\roman*)]
\item ``Idealized'': Segments with high levels of expert agreement.
\item ``Proto'': Segments where approximately half of the experts labeled them
as ``other'', while the other half assigned one of the remaining five labels.
\item ``edge'': Segments where experts were roughly split between two of the
five named patterns.
\end{enumerate*}
Figure~\ref{fig:ex} shows an example of EEG patterns with different level of
agreement.

\begin{figure}[tbp]
\centering
\includegraphics[width=.48\textwidth]{example}
\caption{Examples of EEG patterns with different levels of expert agreement,
sourced from the HMS dataset.}
\label{fig:ex}
\end{figure}

\subsection{Data Preparation}

The data preparation process involves multiple steps to extract spectrogram
features from EEG data, enhance signal quality, and ensure the features are
appropriately scaled and normalized for analysis.

\textbf{Wavelet denoising}: EEG signals were denoised using wavelet
transformation to improve signal quality by decomposing the signal into wavelet
coefficients, applying thresholding to remove noise, and reconstructing the
denoised signal.
%via \pkg{pywt} \proglang{Python} package

\textbf{Feature Extraction}: A bipolar montage technique is applied to the
denoised EEG signals to calculate the difference between adjacent electrode
signals, generating four representative montage features from the original 19
features. Figure~\ref{fig:inion} illustrates the bipolar electrode chains used
in this study.

\textbf{Spectrogram transformation}: The denoised EEG signals are converted
into Mel-spectrograms, which are log-transformed to enhance spectral
information and compress the dynamic range. Figure~\ref{fig:EEG} shows an
example of the denoised EEG signal and its corresponding transformated
spectrogram.
%using the \pkg{librosa} \proglang{Python} package

\textbf{Normalizatio}n: The Mel-spectrogram values are standardized to a range
of $[-1,1]$ to ensure consistent feature scaling and facilitate model training.

\textbf{Imputation}: Missing values in the data are imputed with the mean of
non-missing values within each spectrogram. If all values are missing,
the spectrogram is set to zero to maintain data consistency.

\begin{figure}[tbp]
\centering
\includegraphics[width=.45\textwidth]{inion}
\caption{Chains employed to determine representative montage features via
a bipolar montage, sourced from Kaggle.}
\label{fig:inion}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{EEG_Signal}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{EEG_Spectrogram}
\end{subfigure}
\caption{An example of the denoised EEG signal (top) and its corresponding
	spectrogram representation (bottom).}
\label{fig:EEG}
\end{figure}

\subsection{Loss Function}

The model outputs an array of predicted probabilities for each label class.
To evaluate its performance, the Kullback-Leibler (KL) divergence metric is
employed. This metric is suitable for comparing two probability distributions: 
the ground truth probability distribution $\XX$ and the predicted probability
distribution $\YY$. The KL divergence between these two distributions is
defined as:
\begin{equation*}
D_\mathrm{KL}(\XX \Vert \YY) = \sum \XX \log\left(\frac{\XX}{\YY}\right)
\end{equation*}

The KL divergence offers the advantage of direct applicability as a loss
function during model training, eliminating the need for an alternative loss
function. Its differentiability enables efficient gradient-based optimization
techniques.

\subsection{Learning Rate Scheduler}

To optimize the model's training process, a cosine annealing learning rate
scheduler was employed. This technique ensures a smooth and gradual decrease in
the learning rate, preventing abrupt changes that can destabilize the training
process. By allowing the model to explore a wider parameter space initially and
then refining its solution towards the end, cosine annealing contributes to
better convergence and reduced oscillations in the loss function.

As depicted in Figure~\ref{fig:cos}, the learning rate for both the pre-trained
UNet and the fine-tuned HCRF exhibit a similar pattern. Both start with a
relatively high learning rate, which gradually decreases over time.
This smooth decay helps ensure stable training and optimal convergence.

\begin{figure}
\centering
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{pretrain}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{finetune}
\end{subfigure}
\caption{The learning rate schedulers for the pre-trained UNet (top) and
the fine-tuned HCRF (bottom).}
\label{fig:cos}
\end{figure}
